Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-16 syntaxsemantics.tex
lots of disorganized edits, mostly focused on justifying use of modalities, some comments for fixing in sec 4

A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$. 

@@

% \subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of word-aligned physical memory addresses 
to 64-bit physical memory values. 
Thus our states $\sigma$ include register maps $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $ and
memory maps $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$.
Of particular note, \lstinline|cr3|, the page table register, is included in the machine state.
% As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.

@@

Programs in our logic are instruction sequences \instrs, which are formed by either a basic instruction \iskip, or prefixing an existing instruction
sequence with an additional instruction (\iseq\instr\instrs).
We model (and later, give program logic rules for) instructions for basic register moves, and reading and writing memory.
The latter require page table walks.
% Figure \ref{fig:coq_addr_translation} gives a slightly simplified version of our address translation code in monadic style:
% starting from the root page table address \textsf{rtv}, the virtual address \textsf{w} is either resolved to a physical address or an error is returned.
% The full set of possible errors is not relevant to this paper; our logic is tailored for kernel code which should not fault,
% so our proof rules guarantee no memory failures occur --- that the page table walk for any dereferenced address will succeed.

@@

The most important instructions we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right).
Thus a store to memory is $\textsf{mov}~[r_m]~r_r$, and a load from memory is $\textsf{mov}~r_r~[r_m]$.
% \[
% \begin{array}{l}
%   \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
%   \textsf{mov}~[r_m]~r_r \\
%   \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
%   \textsf{mov}~r_r~[r_m]
%   \end{array}
%   \]
Operationally, each first translates the virtual address stored in register $r_m$ to a physical address,
then either updates the memory at that physical location with the contents of $r_r$ (for store)
or updates $r_r$ with the contents of that memory (for load).
% The semantics of each is realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location
% per the description in Section \ref{sec:backgroundonmachinemodel},
% then either updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
% or loading the value at that physical memory location into the register $r_r$
% (\TirNameStyle{WriteToRegFromMem}). 
Our formalization includes
additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers),
basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.
The semantics model updates to the instruction pointer, though we elide that register (\lstinline|rip|)
from specifications in the paper for clarity.

>>>>

A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$.\footnote{\add{$\reg$ abbreviates \emph{general register}, which includes integer registers (e.g., \lstinline|rax|),
control registers (e.g., \lstinline|cr3|), and segmentation registers (which we do not discuss in detail,
but are still used in a limited way in 64-bit mode on modern x86-64 processors).
}} 

@@

% \subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of word-aligned physical memory addresses 
to 64-bit physical memory values. 
Thus our states $\sigma$ include register maps $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $ and
memory maps $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$
\add{which segment memory into page-sized increments}.
Of particular note, \lstinline|cr3|, the page table register, is included in the machine state.
% As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.

@@

Programs in our logic are instruction sequences \instrs, which are formed by either a basic instruction \iskip, or prefixing an existing instruction
sequence with an additional instruction (\iseq\instr\instrs).
We model (and later, give program logic rules for) instructions for \del{basic} register \replace{moves}{loads and stores}, and reading and writing memory.
The latter require page table walks.
% Figure \ref{fig:coq_addr_translation} gives a slightly simplified version of our address translation code in monadic style:
% starting from the root page table address \textsf{rtv}, the virtual address \textsf{w} is either resolved to a physical address or an error is returned.
% The full set of possible errors is not relevant to this paper; our logic is tailored for kernel code which should not fault,
% so our proof rules guarantee no memory failures occur --- that the page table walk for any dereferenced address will succeed.

@@

The most important instructions we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right).
Thus a store to memory is $\textsf{mov}~[r_m]~r_r$, and a load from memory is $\textsf{mov}~r_r~[r_m]$.
% \[
% \begin{array}{l}
%   \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
%   \textsf{mov}~[r_m]~r_r \\
%   \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
%   \textsf{mov}~r_r~[r_m]
%   \end{array}
%   \]
Operationally, each first translates the virtual address stored in register $r_m$ to a physical address \add{(a page table walk)},
then either updates the memory at that physical location with the contents of $r_r$ (for store)
or updates $r_r$ with the contents of that memory (for load).
% The semantics of each is realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location
% per the description in Section \ref{sec:backgroundonmachinemodel},
% then either updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
% or loading the value at that physical memory location into the register $r_r$
% (\TirNameStyle{WriteToRegFromMem}). 
Our formalization includes
additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers),
basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.
The semantics model updates to the instruction pointer, though we elide that register (\lstinline|rip|)
from specifications in the paper for clarity.
\looseness=-1
